Local PCA Shows How Population Structure Differs Along the Genome

### 2016-06-14 writeup/local_pca_paper.tex
update

Dimension reduction techniques,
such as principal components,
are often used to discover and display large-scale structure in genomic datasets
found in the patterns of kinship
between the genotyped individuals,
and to control for the confounding effects of population structure in genome-wide association studies.
The genome-wide mean kinship this uses
is an average of the relationships across all locus-specific genealogical trees.
However, many biological factors,
including linked selection,
can systematically skew patterns of kinship over intermediate genomic scales.
We show how to use principal components analysis (PCA) to describe this meso-scale variation in kinship,
and apply the method to genomic data from three species.
In each species we find substantial variation on the scale of megabases to tens of megabases.
In a global human dataset, small, discontinuous variation is likely explained by polymorphic chromosomal inversions.
In a dataset of African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection.
In a range-wide dataset of \textit{Medicago truncatula},
common axes of variation in population structure are shared between chromosomes,
correlate with local gene density,
and may be caused by background selection or local adaptation.
The method is a useful addition to the exploratory toolbox
of population genomics,
and is implemented as an R package.

@@

Realized kinship summarizes the shapes
of the genealogical that relate the samples
at each location along the genome.
Since these trees vary along the genome, so does realized kinship;
but averaging over sufficiently many trees we hope to get a stable estimate,
independent of the genomic region chosen.
This hope is not entirely justified: for instance,
kinship on sex chromosomes is expected to differ from the autosomes;
and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
\citep{maynardsmith1974hitchhiking,barton2000genetic,neher2013genetic,charlesworth2012effects},
and therefore patterns of kinship.
Indeed,
chromosome-scale variation in diversity and divergence has been observed in many species
\citep[e.g.][]{langley2012genomic};
species phylogenies can differ along the genome 
due to incomplete lineage sorting \citep[e.g.][]{pease2013accurate},
adaptive introgression and/or local adaptation \citep[e.g.][]{ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
and theoretical expectations predict that geographic patterns of relatedness should depend on selection
\citet{charlesworth2003review}.
Nonetheless, it is not generally known to what extent patterns of kinship vary along the genome,
nor what the major axes of variation are.

Patterns in genome-wide kinship are often summarized
by applying principal components analysis (PCA) \citep{patterson2006population} 
to the genetic covariance matrix,
as pioneered by \citet{menozzi1978synthetic}.
% About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
% Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.  
% PCA results are derived from the covariance matrix of genotype matrix. 
The results of PCA can be directly related to the underlying genealogical history of the samples, 
such as time to most recent common ancestor and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}. 
These patterns are often geographical, producing ``maps'' of population structure
that reflect the samples' geographic origin distorted by rates of gene flow
\citep{novembre2008genes},
although other patterns emerge if recent migration or nongeographic kinship patterns are important
\citep{astle2009population}.
% Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
% Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
% Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}. 
% However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}

@@

% To interpret the results of MDS, we combine known genome feature information for each species, such as the
% distribution of inversions, and heterochromatin and gene density along the genome. 
Each species showed distinct patterns, reflecting differences in their biology;
before presenting results,
it may help to have a prior idea of what factors are expected to affect population structure,
and how.
PCA summarizes patterns in kinship found in the genetic covariance matrix,
which is an average across locus-specific genealogies.
% However, different parts of genome have different genetic features. 
% First, each site of DNA may have different gene tree. 
% The covariance matrix of genotype matrix averages those gene trees. 
If individuals are closer in the genealogies of a given genomic region, 
they tends to be more close in the PCA maps for that region. 
% Different DNA segments may have different gene tree and therefore different population structure for those segments. 
% Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
% Selective sweeps cause local recent ancestry or short trees.
% Background selection causes shallow ones.
Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
A genomic region with many targets for selection experiences
background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
which would tend to generally shorten genealogical trees in the region,
similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
% Balancing selection causes deep trees.
On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
and population structure locally describes which individuals have which alleles
rather than geographical proximity.
Finally, 
since recombination is suppressed between opposite orientations of a chromosomal inversion
near its breakpoints,
genealogies in these regions separate samples carrying the two orientations of the inversion.
In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
(indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
as a way to identify inversions).
% if a chromosomal inversion is polymorphic in the sample, 
% the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints. Recombination suppression across inversions thus results in different genome structure and population structure. 
% Other effects, like noise, introgression might also influence population structure.
We will look for strong variation in population structure shared across large regions of the genome.
Many of the effects listed above (such as single selective sweeps or inversions) 
are not expected to have similar effects on population structure in different regions of the genome
because of randomness in which samples end up in which group;
but if these coincide with a region of reduced recombination (such as an inversion),
these could drive major patterns of variation.
The more subtle effects of genome-wide linked selection could be shared across large regions
due to large-scale variation in gene density
(as by background selection or local adaptation at many genes).

@@

\paragraph{\textit{Drosophila melanogaster}:}
We also used whole-genome sequencing data from Drosophila Population Genomics Project (DPGP) and John Pool's lab \citep{lack2015drosophila}, 
which together has 380 samples from 16 countries across Africa and Europe.
\plr{Check: are all of our samples described in \citet{lack2015drosophila}? If yes, just say DPGP; if no, add appropriate citation (ask John what this should be). Also, add website?}
Since the \textit{Drosophila} samples are from inbred lines, we treat the samples as haploid when recoding;
regions with residual heterozygosity were marked as missing in the original dataset.
Each chromosome arm we investigated (Chr2L, Chr2R, Chr3L, and Chr3R) has 2--3 million SNPs.
Due to high density of missing data for some parts in the genome, 
we first deleted any samples with more than 8\% missing genotypes, 
and then deleted positions with more than 20\% missing data. 
(We chose these cutoffs as the tails of the relevant empirical distributions.)

@@

\plr{I'm talking about flipping signs up here, even though in the code we do it when computing variances. Do you think it makes sense here?}
Since eigenvectors are still only defined up to sign,
when comparing between windows we choose the sign to best match each other:
after choosing $\pcone_1$, for instance, and $u$ is the first eigenvector obtained from the covariance matrix
for window $j$,
then $\pcone_j = \pm u$,
where the sign is chosen according to which of 
$\sum_i (\pcone_{i1} - u_i)$ or
$\sum_i (\pcone_{i1} + u_i)$ 
is smaller.

@@

\begin{figure}
    \begin{center}
       \includegraphics{Fig2_pca_plots_allchr_3peaks_byMDS}
    \end{center}
    \caption{      
        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_chr2L}. 
        Each point represents a sample. 
        The plots are plotting the second principal components against the first principal components. 
        The first column shows the windows colored green in Figure 1; the second column shows orange windows; and third shows purple windows.
         \label{fig:pca_chr2L}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
       \includegraphics{Fig6_MDS_1D_against_gene_count_all_chr_win104_with_lm}
    \end{center}
    \caption{
        First MDS coordinate against gene density for all 8 chromosomes. 
        The first MDS coordinate is significantly correlated with gene count (r=0.146, p=$2.2\times 10^{-16}$). 
        (See supplementary for each single chromosome's MDS result against gene count for \textit{Medicago})
        \label{fig:mds_gene_count}
    }
\end{figure}

>>>>

Dimension reduction techniques,
such as principal component analysis (PCA),
are often used to discover and display large-scale structure in genomic datasets
found in the patterns of kinship
between the genotyped individuals,
and to control for the confounding effects of population structure in genome-wide association studies.
The genome-wide mean kinship that principal component uses
is an average of the relationships across all locus-specific genealogical trees.
However, many biological factors,
including linked selection,
can systematically skew patterns of kinship over intermediate genomic scales.
We show how to use principal components analysis (PCA) to describe this meso-scale variation in kinship,
and apply the method to genomic data from three species.
In each species we find that population structure varies on the scale of megabases to tens of megabases.
In a global human dataset, small, discontinuous variation is likely explained by polymorphic chromosomal inversions.
In a dataset of primarily African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection.
In a range-wide dataset of \textit{Medicago truncatula},
common axes of variation in population structure are shared between chromosomes,
correlate with local gene density,
and may be caused by background selection or local adaptation.
The method is a useful addition to the exploratory toolbox
of population genomics,
and is implemented as an R package.

@@

Realized kinship summarizes the shapes
of the genealogical trees that relate the samples
at each location along the genome.
Since these trees vary along the genome, so does realized kinship;
but averaging over sufficiently many trees we hope to get a stable estimate,
independent of the genomic region chosen.
This hope is not entirely justified: for instance,
kinship on sex chromosomes is expected to differ from the autosomes;
and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
\citep{maynardsmith1974hitchhiking,barton2000genetic,neher2013genetic,charlesworth2012effects},
and therefore patterns of kinship.
Indeed,
chromosome-scale variation in diversity and divergence has been observed in many species
\citep[e.g.][]{langley2012genomic};
species phylogenies can differ along the genome 
due to incomplete lineage sorting \citep[e.g.][]{pease2013accurate},
adaptive introgression and/or local adaptation \citep[e.g.][]{ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
and theoretical expectations predict that geographic patterns of relatedness should depend on selection
\citep{charlesworth2003review}.
Nonetheless, it is not generally known to what extent patterns of kinship vary along the genome,
nor what the major axes of variation are.

Patterns in genome-wide kinship are often summarized
by applying principal components analysis (PCA) \citep{patterson2006population} 
to the genetic covariance matrix,
as pioneered by \citet{menozzi1978synthetic}.
% About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
% Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.  
% PCA results are derived from the covariance matrix of genotype matrix. 
The results of PCA can be directly related to the underlying genealogical history of the samples, 
such as time to most recent common ancestor and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}. 
These patterns are often geographical, producing ``maps'' of population structure
that reflect the samples' geographic origin distorted by rates of gene flow
\citep{novembre2008genes},
although other patterns emerge if recent migration or nongeographic kinship patterns are having a strong role in relatedness determining.
\citep{astle2009population}.
% Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
% Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
% Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}. 
% However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}

@@

% To interpret the results of MDS, we combine known genome feature information for each species, such as the
% distribution of inversions, and heterochromatin and gene density along the genome. 
Each species showed distinct patterns, reflecting differences in their biology;
before presenting results,
it may be helpful to have a prior idea of what factors are expected to affect population structure,
and how.
PCA summarizes patterns in kinship found in the genetic covariance matrix,
which is an average across locus-specific genealogies.
% However, different parts of genome have different genetic features. 
% First, each site of DNA may have different gene tree. 
% The covariance matrix of genotype matrix averages those gene trees. 
If individuals are closer in the genealogies of a given genomic region, 
they tend to be closer in the PCA maps for that region. 
% Different DNA segments may have different gene tree and therefore different population structure for those segments. 
% Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
% Selective sweeps cause local recent ancestry or short trees.
% Background selection causes shallow ones.
Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
A genomic region with many targets for selection experiences
background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
which would tend to shorten genealogical trees in the region,
similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
% Balancing selection causes deep trees.
On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
and population structure locally describes which individuals have which alleles
rather than geographical proximity.
Finally, 
since recombination is suppressed between opposite orientations of a chromosomal inversion
near its breakpoints,
genealogies in these regions separate samples carrying the two orientations of the inversion.
In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
(indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
as a way to identify inversions).
% if a chromosomal inversion is polymorphic in the sample, 
% the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints. Recombination suppression across inversions thus results in different genome structure and population structure. 
% Other effects, like noise, introgression might also influence population structure.
We will look for strong variation in population structure shared across large regions of the genome.
Many of the effects listed above (such as single selective sweeps or inversions) 
are not expected to have similar effects on population structure in different regions of the genome
because of randomness in which samples end up in which group;
but if these coincide with a region of reduced recombination (such as an inversion),
these could drive major patterns of variation.
The more subtle effects of genome-wide linked selection could be shared across large regions
due to large-scale variation in gene density
(as by background selection or local adaptation at many genes).

@@

\paragraph{\textit{Drosophila melanogaster}:}
We also used whole-genome sequencing data from Drosophila Population Genomics Project (DPGP) \citep{lack2015drosophila}, 
which together has 380 samples from 16 countries across Africa and Europe.
\plr{Check: are all of our samples described in \citet{lack2015drosophila}? If yes, just say DPGP; if no, add appropriate citation (ask John what this should be). Also, add website?}
Since the \textit{Drosophila} samples are from inbred lines, we treat the samples as haploid when recoding;
regions with residual heterozygosity were marked as missing in the original dataset.
Each chromosome arm we investigated (Chr2L, Chr2R, Chr3L, and Chr3R) has 2--3 million SNPs.
Due to high density of missing data for some parts in the genome, 
we first deleted any samples with more than 8\% missing genotypes, 
and then deleted positions with more than 20\% missing data. 
(We chose these cutoffs as the tails of the relevant empirical distributions.)

@@

\plr{I'm talking about flipping signs up here, even though in the code we do it when computing variances. Do you think it makes sense here?}
Since eigenvectors are still only defined up to sign,
when comparing between windows we choose the sign to best match each other:
after choosing $\pcone_1$, for instance, and $u$ is the first eigenvector obtained from the covariance matrix
for window $j$,
then $\pcone_j = \pm u$,
where the sign is chosen according to which of 
$\sum_i (\pcone_{i1} - u_i)^{2}$ or
$\sum_i (\pcone_{i1} + u_i)^{2}$ 
is smaller.

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=1\textwidth]{Fig2_pca_plots_allchr_3peaks_label_update}
    \end{center}
    \caption{      
        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_chr2L}. 
        Each point represents a sample. 
        The plots are plotting the second principal components against the first principal components. 
        The first column shows the windows colored green in Figure 1; the second column shows orange windows; and third shows purple windows.
         \label{fig:pca_chr2L}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{MDS_1D_against_gene_count_all_chr_win104_with_lm}
    \end{center}
    \caption{
        First MDS coordinate against gene density for all 8 chromosomes. 
        The first MDS coordinate is significantly correlated with gene count (r=0.146, p=$2.2\times 10^{-16}$). 
        (See supplementary for each single chromosome's MDS result against gene count for \textit{Medicago})
        \label{fig:mds_gene_count}
    }
\end{figure}
